---
title: "Assignment 10: Data Scraping"
author: "Shiqi Zheng"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)
library(here)
here()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
LWSP <- read_html(
  'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')
LWSP
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
name <- LWSP %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
name

PWSID <- LWSP %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
PWSID

ownership <- LWSP %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership

Maximum.Day.Use <- LWSP %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
Maximum.Day.Use

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 
df <- data.frame(
  "WaterSystemName" = rep(name,12), 
  "PWSID" = rep(PWSID, 12),
  "Ownership" = rep(ownership,12),
  "MaximumDayUse" = as.numeric(Maximum.Day.Use),
  "Month" = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul",
              "Aug", "Sep", "Oct", "Nov", "Dec"),
  "Year" = rep(2022,12)
) %>% 
  mutate(Date = my(paste(Month,"-",Year)))
head(df)

# ?? a row; modify reflect year

#5 
ggplot(df,aes(x=Date,y=MaximumDayUse)) + 
  geom_line() + 
   scale_x_date(date_labels = "%b", date_breaks = "1 month") + 
  labs(title = "Water Maximum Daily Withdrawals for Durham in 2022",
       y="Maximum Day Use (MGD)",
       x="Date")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
scrape.it <- function(year, PWSID){
   LWSP <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', 
                                  PWSID, '&year=',year))
#scape data   
name <- LWSP %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()
PWSID <- LWSP %>% html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% html_text()
ownership <- LWSP %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()
Maximum.Day.Use <- LWSP %>% html_nodes("th~ td+ td") %>% html_text()

# create dataframe
df <- data.frame(
  "WaterSystemName" = rep(name,12), 
  "PWSID" = rep(PWSID, 12),
  "Ownership" = rep(ownership,12),
  "MaximumDayUse" = as.numeric(Maximum.Day.Use),
  "Month" = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", 
              "Aug", "Sep", "Oct", "Nov", "Dec"),
  "Year" = rep(year,12)
) %>% 
  mutate(Date = my(paste(Month,"-",Year)))

return(df)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
the_df <- scrape.it(2015,'03-32-010')
ggplot(the_df,aes(x=Date,y=MaximumDayUse)) + 
  geom_line() + 
  scale_x_date(date_labels = "%b", date_breaks = "1 month") + 
  labs(title = "Water Maximum Daily Withdrawals for Durham in 2015",
       y="Maximum Day Use (MGD)",
       x="Date")
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 
Asheville_df <- scrape.it(2015,'01-11-010')
combine_df <- rbind(Asheville_df,the_df)
ggplot(combine_df,aes(x=Date,y=MaximumDayUse,color = WaterSystemName)) + 
  geom_line() + 
  scale_x_date(date_labels = "%b", date_breaks = "1 month") + 
  labs(title = "Water Maximum Daily Withdrawals for Durham and Asheville in 2015",
       y="Maximum Day Use (MGD)",
       x="Date",
       color ="System Name")

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data}
#9 
the_years <- seq(2010,2021)
the_PWSID <-  rep('01-11-010',length(the_years))

dfs_Asheville <- map2(the_years, the_PWSID , scrape.it) %>% bind_rows()
ggplot(dfs_Asheville, aes(x = Date, y = MaximumDayUse)) +
  geom_line() +
  geom_smooth(method = 'loess', se = FALSE) +
  labs(title = "Asheville's Maximum Daily Withdrawals from 2010 to 2021",
       y = "Maximum Day Use (MGD)",
       x = "Date")
```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?

> Answer: From 2010 to 2015, Asheville has a decreasing trend in water usage. From 2015 to 2020, Asheville has a increasing trend in water usage through time.
